Coherent Hyper-Raman Four Wave Mixing Vibrational Spectroscopy

### 2024-07-22 paper.tex

\begin{abstract}
Nonlinear, four wave mixing vibrational spectroscopies are often used to probe intramolecular vibrational coupling and relaxation dynamics in isotropic media.
Three wave mixing vibrational spectroscopies are similarly used to understand the spectroscopy and dynamics of interfacial species.
Most of these methods rely on infrared or Raman transitions to generate output. 
Implementing a nonlinear spectroscopy involving hyper-Raman transitions, first appearing under third order perturbation theory, can provide a useful analogue to infrared and Raman based nonlinear spectroscopies to understand the spectroscopy and dynamics of isotropic systems.
Hyper Difference Frequency Generation (Hyper-DFG, or HDFG) spectroscopy, also known as singly vibrationally enhanced spectroscopy (SIVE), is an underdeveloped, hyper-Raman based four wave mixing vibrational spectroscopy. 
Despite several experimental reports on HDFG, its spectroscopic properties have not been fully explored.
To this end, we explore the selection rules and response functions of HDFG spectroscopy.
Through the hyper-Raman $A,B,C$ terms, we explore the selection rules of singly and doubly resonant HDFG spectroscopy, and show HDFG is a hybrid IR-hyper-Raman technique.
Since all infrared active modes have non-zero hyper-Raman activity, this makes HDFG a type of upconverted infrared spectroscopy.
HDFG is found to be uniquely sensitive to non-Condon effects in centrosymmetric species.
Through a simple comparison test, HDFG output is shown to be similar in magnitude to vibrational sum frequency generation (vSFG).
Since HDFG can also be implemented as a two beam experiment, similar to vSFG, this makes HDFG a spectroscopy feasible for practitioners of vSFG.
Through a simple treatment of orientational averaging, HDFG provides a simple method to extract hyper-Raman hyperpolarizabilities ($\beta$) of infrared active vibrations.
HDFG shows promise as a method to disentangle vibrational spectra and dynamics in isotropic systems without need for anharmonicities.

@@

\section{Introduction}
Coherent multidimensional spectroscopy (CMDS) is a family of three and four wave mixing methods which form the optical analogue of multidimensional nuclear magnetic resonance (NMR) spectroscopy.\cite{RN103, Cho2008}
Multiresonant, four wave mixing CMDS experiments, first proposed by Oudar and Shen in 1980,\cite{RN307} directly probe coupling and correlations between different vibrational, electronic, and vibronic states. \cite{RN307, RN281, RN103, RN342, Cho2008, Dong2015, Ogilvie2019, Bonn2021, RN325} 
CMDS has resolved anharmonicities, ultrafast dynamics, and other intramolecular couplings in numerous systems. \cite{RN342, RN343, RN324, Czech2015, Gaynor2017, Ogilvie2019, RN325}

Not all CMDS methods are used to dissect intramolecular couplings.\cite{Shen1987_CPL}
The most well known example of such a spectroscopy is vibrational sum frequency generation (vSFG), a three wave mixing technique used to probe vibrational structure and dynamics at interfaces. \cite{Shen1987_CPL, RN224}
Non-parametric, hyper difference frequency generation (hyper-DFG, or HDFG) spectroscopy, similar to difference frequency generation (\autoref{fig:comparisonwmel}), is a four wave mixing method which, when detuned from  electronic resonances, does not probe intramolecular couplings. 
HDFG was the first infrared CMDS method to successfully discriminate against non-resonant background.\cite{RN351, RN352}
HDFG has also been referred to as singly vibrationally enhanced (SIVE) spectroscopy. \cite{RN351}
In HDFG, an infrared pulse is resonant with a vibrational mode, and two other input pulses are used to stimulate a scattering process.
HDFG was documented long ago to have characteristics similar to both infrared spectroscopy and spontaneous hyper-Raman scattering. \cite{RN352}
Hyper-Raman scattering is the two photon analogue to Raman scattering. \cite{Cyvin1965, Terhune1965, Andrews1978}
Hyper-Raman scattering cross sections are commonly several orders of magnitude smaller than Raman scattering cross sections, making hyper-Raman a difficult process to detect.\cite{RN515, Kelley2010} 
However, unlike Raman spectroscopy, all infrared active modes are hyper-Raman active, which allows infrared active modes to be probed through an inelastic scattering event. \cite{Andrews1978}
The different selection rules of hyper-Raman scattering make it a unique alternative to Raman scattering to understand vibrational spectra and vibronic coupling in molecular systems. \cite{Olson2018}
Some years ago, Cho et al. proposed parametric six wave mixing techniques as the coherent analogue of hyper-Raman scattering. \cite{Cho1997, Cho1998}
However, since most six wave mixing methods cascade into four wave mixing processes and their homodyned output would be dependent upon the hyper-Raman scattering cross section, \cite{RN243, RN515, Cho1997, Cho2000_Cascade} it is preferable to investigate four wave mixing analogues to hyper-Raman scattering.

\begin{figure*}[!htbp]
	\centering
	\includegraphics[width=6.66in]{comparisonwmel.png}
	\caption{Wave Mixing Energy Level (WMEL) diagrams of difference frequency generation (DFG), spontaneous Raman scattering, hyper difference frequency generation (HDFG) and spontaneous hyper-Raman scattering relevant to this discussion. \cite{RN286}
	Solid and dashed horizontal lines indicate real and virtual states, whereas solid and dotted arrows indicate ket and bra side transitions, respectively. 
	}
	\label{fig:comparisonwmel}
\end{figure*}

@@

An experimental complexity in executing Raman based four wave mixing experiments is the demand of three separate input pulses, often at different frequencies.
For example, in the case of DOVE, two of the three inputs must be scanable infrared pulses. \cite{RN345} 
Only recently have methods been developed for seamless tuning and scanning of ultrafast optical parametric amplifiers across swaths of frequency space. \cite{RN162, McDonnell2024, SkyeOPA, KyleOPA}
Since HDFG demands only a single resonance to be scanned, the experiment can be performed using two input pulses: one tunable, infrared OPA to scan across vibrational resonances, and a two photon absorption from the signal process of a separate OPA, or output from the oscillator which pumps the infrared OPA. \cite{Wang2021}
As such, laboratories experienced in sum frequency generation (SFG) spectroscopy can perform HDFG spectroscopy to understand bulk dynamics and spectroscopy using nearly the same setup.\cite{Shen1987_CPL}

Inspired by recent work which demonstrated the presence of HDFG in ultrafast DOVE and terahertz-IR-visible (TIRV) experiments, we investigate the parameters which result in both SR and DR-SIVE output. \cite{Cho2000, Bonn2024, McDonnell2024}
This paper is organized as follows.
In \autoref{steadystate}, we identify the selection rules that drive singly and doubly resonant HDFG, first through the Placzek approximation, and then through hyper-Raman $A,B,C$ coefficients.
Quantitative aspects of HDFG: a comparison of vSFG and HDFG output, a method to extract hyper-Raman hyperpolarizabilities ($\beta_{ijk}$) from HDFG spectra, are discussed in \autoref{quant}.
Mixed time-frequency domain HDFG is discussed in \autoref{mixeddomain}.

@@

It is useful to expose relationships between transition dipoles, Raman polarizabilities and hyper-Raman hyperpolarizabilities in the driven limit. \cite{Druet1978, Simpson2004}
In the driven limit, under the electric dipole approximation, the I$^{th}$ component of the third order nonlinear output polarization, ${P}^{(3)}_I$, of any four wave mixing process, induced by electric fields E, at output frequency $\omega_4$ is written as (using Einstein summation) \cite{RN307}
\begin{equation} \label{polarization}
{P}^{(3)}_I (\omega_4)  = \chi^{(3)}_{IJKL} E_J E_K E_L 
\end{equation}
where $\chi^{(3)}_{IJKL}$ is the IJKL element of the third order electrical susceptibility, a rank four tensor, generally written as
\begin{equation}
	\chi^{(3)}_{IJKL} = NF(\omega_4) \langle \gamma_{ijkl} \rangle
\end{equation}
where N is a number density, F is the Lorentz local field factor, and $\gamma_{ijkl}$ is the third order polarizability (i.e., second hyperpolarizability). 
The brackets indicate an orientational average. 
Uppercase letters refer to laboratory frame coordinates and lower case letters refer to molecular frame coordinates.

@@

\subsection{Placzek Method}
We first investigate the selection rules through a Placzek type formalism.
Contracting over the virtual states forms the hyper-Raman hyperpolarizability $\beta$.\cite{Long1970} 
Assuming $\omega_2$, $\omega_3$ are significantly detuned from any resonance (\autoref{fig:comparisonwmel}),\cite{Placzek1934, Long1970, Altmann1982} \autoref{sivegamma} is written as 
\begin{equation}\label{sivebeta}
	\gamma_{ijkl}^{vg} =	-\frac{1}{\varepsilon_0} \frac{1}{4D \hbar}\frac{\beta^{vg}_{ijk} \mu^{gv}_{l}}{\Delta_{gv}} \rho_{gg}
\end{equation}
All infrared active transitions are hyper-Raman active, making HDFG allowed for any infrared active transition. \cite{Andrews1978}
It is important to note that the symmetry properties of $\beta_{ijk}$ change dependent upon whether $\omega_2$ and $\omega_3$ are degenerate. \cite{Denisov1986, Kozich2007}
Taylor expanding the dipole and first hyperpolarizability operators in terms of an n$^{\text{th}}$ reduced normal mode coordinate $Q_n$ about equilibrium as\cite{Long1970, Shen90}
\begin{subequations}
	\begin{equation}
		\mu_l = \mu_{l,0} + \left(\frac{\partial \mu_l}{\partial Q_n}\right)_0 Q_n + \order{{Q_m Q_n}}
	\end{equation}
	\begin{equation}
		\beta_{ijk} = \beta_{ijk,0} + \left(\frac{\partial \beta_{ijk}}{\partial Q_n}\right)_0 Q_n + \order{{Q_m Q_n}}
	\end{equation}
\end{subequations}
and substituting into \autoref{sivebeta} gives the HDFG hyperpolarizability to $\order{Q_n}$ as \begin{equation}\label{SIVEselection}
	\gamma_{ijkl}^{vg} =	-\frac{1}{\varepsilon_0} \frac{1}{8D m_n \omega_{vg}}  \frac{1}{{\Delta_{gv}}} \ \left(\frac{\partial \beta^{vg}_{ijk}}{\partial Q_n}\right)_0 \left({\frac{\partial \mu^{gv}_{l}}{\partial Q_n}}\right)_0  \rho_{gg}
\end{equation}
where $\mel{v}{Q_n}{g} = \sqrt{\frac{\hbar}{2\omega_{vg}}}$.  \cite{RN230}
Since this expression is non-zero in the harmonic oscillator limit, HDFG output is allowed for harmonic transitions. 
Similar to SFG, HDFG does not demand anharmonic ground state potential energy surfaces for spectral output. \cite{Shen94, Cho2000}
This selection rule is generally valid for any HDFG or hyper sum frequency generation (HSFG) process in the $-\vec{k}_1 + \vec{k}_2  + \vec{k}_3$ and $\vec{k}_1 + \vec{k}_2  + \vec{k}_3$ phasematching geometries when $\omega_2$ and $\omega_3$ are sufficiently detuned from any resonance.

\begin{figure}[!htbp]
	\centering
	\includegraphics[width=3.375in]{hdfg.png}
	\caption{WMEL diagrams of (a) singly resonant and (b) doubly resonant HDFG. 
	}
	\label{fig:hdfg}
\end{figure}
\subsection{Albrecht Type Expansion}
While Placzek type arguments provide the general source of HDFG output, it is insightful to interpret the origin of signal in terms of $A,B,C$ terms, analogous to those found in Albrecht's treatment of Raman spectroscopy.\cite{Albrecht1961, Ziegler1988} 
We write $\beta_{ijk}$ in \autoref{SIVEselection} using the hyper-Raman ABC formalism introduced by Chung and Ziegler to understand how vibronic coupling manipulates selection rules for singly and doubly resonant HDFG (\autoref{fig:hdfg}). \cite{Ziegler1988, Baranov1990}
To employ this formalism, we write the states in terms of a Born-Oppenheimer basis $\ket{a,b}$, where $\ket{a,b} = |a(Q)) \otimes \ket{b}$ for electronic states $\{|a(Q))\}$ and vibrational states $\{\ket{b}\}$ (i.e., adiabatic approximation). \cite{BornOppenheimer, Tang1970}
The dependence of the electronic states on the adiabatic parameter $Q$ is suppressed for simplicity.
The states are labeled as shown in \autoref{fig:hdfg}.
For consistency with previous reports, we use $\vec{R}$ to denote electric transition dipole moments ($\vec{\mu}$ is reserved for transitions on the ground electronic state). \cite{Tang1970}
Following the approach which gave \autoref{sivegamma}, we find
\begin{equation}\label{drgamma_notaylor}
	\gamma_{ijkl} = -\frac{1}{\varepsilon_0} \frac{1}{4D \hbar^3} \sum_{m,n,e,v'} \frac{
		R_{i}^{gv, ev'} 
		R_{j}^{ev',mn} 
		R_{k}^{mn,g0} 
		R_{l}^{g0,gv} 
	}{\Delta_{g0,gv}
		\Delta_{ev', mn}
		\Delta_{mn, g0}
	}
\end{equation}
where $R_{i}^{ab,cd}$ is the i$^{th}$ element of $\mel{a,b}{\vec{R}}{c,d}$.
It is common in the hyper-Raman community to write $\Delta_{ev', mn} \Delta_{mn, g0} = \Delta_{ev', g0}$.
Using our definition of vibronic states, we write $\vec{M}^{ab} = (a|\vec{M}|b)$ so that, for example,
$R_{i}^{gv,ev'} = \mel{v}{M_i^{ge}}{v'}$.
Using these ideas rewrites \autoref{drgamma_notaylor} as
\begin{equation}\label{drgamma_notaylor_1}
	\begin{split}
		\gamma_{ijkl} &= -\frac{1}{\varepsilon_0} \frac{1}{4D \hbar^3} \sum_{m,n,e,v'} \frac{
			\mel{v}{M_{i}^{ge}}{v'} 
			\mel{v'}{M_{j}^{em}}{n}
			\mel{n}{M_{k}^{mg} }{v}
			\mel{v}{\mu_{l}}{0}
		}{\Delta_{g0,gv}
			\Delta_{ev',g0}	} \\
	\end{split}
\end{equation}
Chung and Ziegler showed some years ago that expanding $\vec{M}^{ij}$ to $\order{Q}$ as
$\vec{M}^{ij} = \vec{M}^{ij}_0 + \sum_z \frac{\partial\vec{M}^{ij}}{\partial Q_z} Q_z$
yields $A, B, C$ coefficients similar to those in the Albrecht formalism of Raman scattering, \cite{Albrecht1961, Ziegler1988} so that
\begin{equation}
		\gamma_{ijkl} \sim \frac{(A_{ijk} + B_{ijk} + C_{ijk})\mel{v}{\mu_{l}}{0}} {\Delta_{g0,gv}}
\end{equation}
The $A$ term depends upon $\order{Q^0}$ transitions (i.e., Condon approximation), the $B$ term depends upon $\order{Q}$ transitions, and the $C$ term depends on $\order{Q^2}$ transitions. 
The C term is suppressed in the following discussion as it depends on one and two photon forbidden transitions. \cite{Ziegler1988, Neddersen1989, Bonang1992}
Note that some reports obtain $A, B, C$ coefficients through a Herzberg-Teller expansion of the electronic states to explicitly expose vibronic couplings in terms of $\partial H / \partial Q $, where H is the electronic Hamiltonian.\cite{HerzbergTeller1933, Petrov1985, Neddersen1989, Baranov1990}
This approach is not used here as the expansion of $\vec{M}^{ij}$ in normal mode coordinates provides sufficient physical insight into hyper-Raman selection rules.

>>>>

\begin{abstract}
Nonlinear, four wave mixing vibrational spectroscopies are often used to probe intramolecular vibrational coupling and relaxation dynamics in isotropic media.
Three wave mixing vibrational spectroscopies are similarly used to understand the spectroscopy and dynamics of interfacial species.
Most of these methods rely on infrared or Raman transitions to generate output. 
Implementing a nonlinear spectroscopy involving hyper-Raman transitions, first appearing under third order perturbation theory, can provide a useful analogue to infrared and Raman based nonlinear spectroscopies to understand the spectroscopy and dynamics of isotropic systems.
Hyper Difference Frequency Generation (Hyper-DFG, or HDFG) spectroscopy, also known as singly vibrationally enhanced spectroscopy (SIVE), is an underdeveloped, hyper-Raman based four wave mixing vibrational spectroscopy. 
Despite several experimental reports on HDFG, its spectroscopic properties have not been fully explored.
To this end, we explore the selection rules and response functions of HDFG spectroscopy.
Through the hyper-Raman $A,B$ terms, we explore the selection rules of singly and doubly resonant HDFG spectroscopy, and show HDFG is a hybrid IR-hyper-Raman technique.
Since all infrared active modes have non-zero hyper-Raman activity, this makes HDFG a type of upconverted infrared spectroscopy.
HDFG is found to be uniquely sensitive to non-Condon effects in centrosymmetric species.
Through a simple comparison test, HDFG output is shown to be similar in magnitude to vibrational sum frequency generation (vSFG).
Since HDFG can also be implemented as a two beam experiment, similar to vSFG, this makes HDFG a spectroscopy feasible for practitioners of vSFG.
Through a simple treatment of orientational averaging, HDFG provides a simple method to extract hyper-Raman hyperpolarizabilities ($\beta$) of infrared active vibrations.
HDFG shows promise as a method to disentangle vibrational spectra and dynamics in isotropic systems without need for anharmonicities.

@@

\section{Introduction}
Coherent multidimensional spectroscopy (CMDS) is a family of three and four wave mixing methods which form the optical analogue of multidimensional nuclear magnetic resonance (NMR) spectroscopy.\cite{RN103, Cho2008}
Multiresonant, four wave mixing CMDS experiments, first proposed by Oudar and Shen in 1980,\cite{RN307} directly probe coupling and correlations between different vibrational, electronic, and vibronic states. \cite{RN307, RN281, RN103, Cho2008} 
Four wave mixing CMDS has resolved anharmonicities, ultrafast dynamics, and other intramolecular couplings in numerous systems. \cite{Cho2008, Gaynor2017, Ziegler2018, Ogilvie2019, Bonn2021, RN325}

Not all CMDS methods are used to dissect intramolecular couplings.\cite{Shen1987_CPL}
The most well known example of such a spectroscopy is vibrational sum frequency generation (vSFG), a three wave mixing technique used to probe vibrational structure and dynamics at interfaces. \cite{Shen1987_CPL, RN224}
Non-parametric, hyper difference frequency generation (hyper-DFG, or HDFG) spectroscopy, similar to difference frequency generation (\autoref{fig:comparisonwmel}), is a four wave mixing method which, when detuned from  electronic resonances, does not probe intramolecular couplings. 
HDFG was one of the first infrared CMDS methods to successfully discriminate against non-resonant background.\cite{RN351, RN352}
HDFG has also been referred to as singly vibrationally enhanced (SIVE) spectroscopy. \cite{RN351}
In HDFG, an infrared pulse is resonant with a vibrational mode, and two other input pulses are used to stimulate a scattering process.
HDFG was documented long ago to have characteristics similar to both infrared spectroscopy and spontaneous hyper-Raman scattering. \cite{RN352}
Hyper-Raman scattering is the two photon analogue to Raman scattering. \cite{Terhune1965, Cyvin1965, Andrews1978}
Hyper-Raman scattering cross sections are commonly several orders of magnitude smaller than Raman scattering cross sections, making hyper-Raman a difficult process to detect.\cite{RN515, Kelley2010} 
However, unlike Raman spectroscopy, all infrared active modes are hyper-Raman active, which allows infrared active modes to be probed through an inelastic scattering event. \cite{Andrews1978}
The different selection rules of hyper-Raman scattering make it a unique alternative to Raman scattering to understand vibrational spectra, electronic structure and vibronic coupling in molecular systems. \cite{Olson2018}
Cho et al. have proposed parametric six wave mixing techniques as the coherent analogue of hyper-Raman scattering. \cite{Cho1997, Cho1998}
However, since most six wave mixing methods cascade into four wave mixing processes and their homodyned output would be dependent upon the hyper-Raman scattering cross section, \cite{RN243,Cho1997, Cho1998, Cho2000_Cascade} it is preferable to investigate four wave mixing analogues to hyper-Raman scattering.

\begin{figure*}[!htbp]
	\centering
	\includegraphics[width=6.66in]{comparisonwmel.png}
	\caption{Wave Mixing Energy Level (WMEL) diagrams of difference frequency generation (DFG), sum frequency generation (SFG), spontaneous Raman scattering, hyper difference frequency generation (HDFG), hyper sum frequency generation (HSFG) and spontaneous hyper-Raman scattering relevant to this discussion. \cite{RN286}
	Solid and dashed horizontal lines indicate real and virtual states, whereas solid and dotted arrows indicate ket and bra side transitions, respectively. 
	}
	\label{fig:comparisonwmel}
\end{figure*}

@@

An experimental complexity in executing Raman based four wave mixing experiments is the demand of three separate input pulses, often at different frequencies.
For example, in the case of DOVE, two of the three inputs must be tunable infrared pulses. \cite{RN345} 
Only recently have methods been developed for seamless tuning and scanning of ultrafast optical parametric amplifiers across swaths of frequency space. \cite{RN162, McDonnell2024}
Since HDFG demands only a single resonance to be scanned, the experiment can be performed using two input pulses: one tunable, infrared OPA to scan across vibrational resonances, and a two photon absorption from the signal process of a separate OPA, or output from the oscillator which pumps the infrared OPA. \cite{Wang2021}
As such, laboratories experienced in sum frequency generation (SFG) spectroscopy can perform HDFG spectroscopy to understand bulk dynamics and spectroscopy using nearly the same setup.

Inspired by recent work which demonstrated the presence of HDFG in ultrafast DOVE and terahertz-IR-visible (TIRV) experiments, we investigate the parameters which result in both SR and DR-SIVE output. \cite{Cho2000, Bonn2024, McDonnell2024}
This paper is organized as follows.
In \autoref{steadystate}, we identify the selection rules that drive singly and doubly resonant HDFG, first through the Placzek approximation, and then through hyper-Raman $A,B,C$ coefficients.
After developing selection rules, some quantitative aspects of HDFG, particularly a comparison of vSFG and HDFG output and a method to extract hyper-Raman hyperpolarizabilities ($\beta_{ijk}$) from HDFG spectra, are discussed in \autoref{quant}.
Finally, mixed time-frequency domain HDFG is discussed in \autoref{mixeddomain}.

@@

It is useful to expose relationships between transition dipoles, Raman polarizabilities and hyper-Raman hyperpolarizabilities in the driven limit. \cite{Simpson2004}
In the driven limit, under the electric dipole approximation, the I$^{th}$ component of the third order nonlinear output polarization, ${P}^{(3)}_I$, of any four wave mixing process, induced by electric fields E, at output frequency $\omega_4$ is written as (using Einstein summation) \cite{RN307}
\begin{equation} \label{polarization}
{P}^{(3)}_I (\omega_4)  = \chi^{(3)}_{IJKL} E_J E_K E_L 
\end{equation}
where $\chi^{(3)}_{IJKL}$ is the IJKL element of the third order electrical susceptibility, a rank four tensor, generally written as
\begin{equation}
	\chi^{(3)}_{IJKL} = NF(\omega_4) \langle \gamma_{ijkl} \rangle
\end{equation}
where N is a number density, F is the Lorentz local field factor, and $\gamma_{ijkl}$ is the third order polarizability (i.e., second hyperpolarizability). 
The brackets indicate an orientational average. 
Uppercase letters refer to laboratory frame coordinates and lower case letters refer to molecular frame coordinates.

@@

\subsection{Placzek Method}
We first investigate the selection rules through a Placzek type formalism.
Contracting over the virtual states forms the hyper-Raman hyperpolarizability $\beta$.\cite{Long1970} 
Assuming $\omega_2$, $\omega_3$ are significantly detuned from any resonance (\autoref{fig:comparisonwmel}),\cite{Placzek1934, Long1970, Altmann1982} \autoref{sivegamma} is written as 
\begin{equation}\label{sivebeta}
	\gamma_{ijkl}^{vg} =	-\frac{1}{\varepsilon_0} \frac{1}{4D \hbar}\frac{\beta^{vg}_{ijk} \mu^{gv}_{l}}{\Delta_{gv}} \rho_{gg}
\end{equation}
All infrared active transitions are hyper-Raman active, making HDFG allowed for any infrared active transition. \cite{Andrews1978}
It is important to note that the symmetry properties of $\beta_{ijk}$ change dependent upon whether $\omega_2$ and $\omega_3$ are degenerate. \cite{Denisov1986, Kozich2007}
Taylor expanding the dipole and first hyperpolarizability operators in terms of an n$^{\text{th}}$ reduced normal mode coordinate $Q_n$ to $\order{Q_n}$ about equilibrium as\cite{Long1970, Shen90}
\begin{subequations}
	\begin{equation}
		\mu_l = \mu_{l,0} + \left(\frac{\partial \mu_l}{\partial Q_n}\right)_0 Q_n 
	\end{equation}
	\begin{equation}
		\beta_{ijk} = \beta_{ijk,0} + \left(\frac{\partial \beta_{ijk}}{\partial Q_n}\right)_0 Q_n
	\end{equation}
\end{subequations}
and substituting into \autoref{sivebeta} gives the HDFG hyperpolarizability to $\order{Q_n}$ as \begin{equation}\label{SIVEselection}
	\gamma_{ijkl}^{vg} =	-\frac{1}{\varepsilon_0} \frac{1}{8D m_n \omega_{vg}}  \frac{1}{{\Delta_{gv}}} \ \left(\frac{\partial \beta^{vg}_{ijk}}{\partial Q_n}\right)_0 \left({\frac{\partial \mu^{gv}_{l}}{\partial Q_n}}\right)_0  \rho_{gg}
\end{equation}
where $\mel{v}{Q_n}{g} = \sqrt{\frac{\hbar}{2\omega_{vg}}}$ if $v-g = \pm 1$.  \cite{RN230}
Since this expression is non-zero in the harmonic oscillator limit, HDFG output is allowed for harmonic transitions. 
Similar to SFG, HDFG does not demand anharmonic ground state potential energy surfaces for spectral output. \cite{Shen94, Cho2000}
This selection rule is generally valid for any HDFG or hyper sum frequency generation (HSFG) process in the $-\vec{k}_1 + \vec{k}_2  + \vec{k}_3$ and $\vec{k}_1 + \vec{k}_2  + \vec{k}_3$ phasematching geometries when $\omega_2$ and $\omega_3$ are sufficiently detuned from any resonance.

\begin{figure}[!htbp]
	\centering
	\includegraphics[width=3.375in]{hdfg.png}
	\caption{WMEL diagrams of (a) singly resonant and (b) doubly resonant HDFG. 
	}
	\label{fig:hdfg}
\end{figure}
\subsection{Albrecht Type Expansion}
While Placzek type arguments provide the general source of HDFG output, it is insightful to interpret the origin of signal in terms of $A,B,C$ terms, analogous to those found in Albrecht's treatment of Raman spectroscopy.\cite{Albrecht1961, Ziegler1988} 
We write $\beta_{ijk}$ in \autoref{SIVEselection} using the hyper-Raman ABC formalism introduced by Chung and Ziegler to understand how vibronic coupling manipulates selection rules for singly and doubly resonant HDFG (\autoref{fig:hdfg}). \cite{Ziegler1988}
To employ this formalism, we write the states in terms of a Born-Oppenheimer basis $\ket{a,b}$, where $\ket{a,b} = |a(Q)) \otimes \ket{b}$ for electronic states $\{|a(Q))\}$ and vibrational states $\{\ket{b}\}$ (i.e., adiabatic approximation). \cite{BornOppenheimer, Tang1970}
The dependence of the electronic states on the adiabatic parameter $Q$ is suppressed for simplicity.
The states are labeled as shown in \autoref{fig:hdfg}.
For consistency with previous reports, we use $\vec{R}$ to denote electric transition dipole moments; $\vec{\mu}$ is reserved for transitions on the ground electronic state. \cite{Tang1970}
Following the approach which gave \autoref{sivegamma}, we find
\begin{equation}\label{drgamma_notaylor}
	\gamma_{ijkl} = -\frac{1}{\varepsilon_0} \frac{1}{4D \hbar^3} \sum_{m,n,e,v'} \frac{
		R_{i}^{gv, ev'} 
		R_{j}^{ev',mn} 
		R_{k}^{mn,g0} 
		R_{l}^{g0,gv} 
	}{\Delta_{g0,gv}
		\Delta_{ev', mn}
		\Delta_{mn, g0}
	}
\end{equation}
where $R_{i}^{ab,cd}$ is the i$^{th}$ element of $\mel{a,b}{\vec{R}}{c,d}$.
It is common in the hyper-Raman community to write $\Delta_{ev', mn} \Delta_{mn, g0} = \Delta_{ev', g0}$.
Using our definition of vibronic states, we write $\vec{M}^{ab} = (a|\vec{M}|b)$ so that, for example,
$R_{i}^{gv,ev'} = \mel{v}{M_i^{ge}}{v'}$.
Chung and Ziegler showed some years ago that expanding $\vec{M}^{ij}$ to $\order{Q}$ \textit{about the equilibrium point of the ground state potential surface} as
$\vec{M}^{ij} = \vec{M}^{ij}_0 + \sum_z \frac{\partial\vec{M}^{ij}}{\partial Q_z} Q_z$
yields $A, B, C$ coefficients similar to those in the Albrecht formalism of Raman scattering, \cite{Albrecht1961, Ziegler1988} so that
\begin{equation}
		\gamma_{ijkl} \sim \frac{(A_{ijk} + B_{ijk} + C_{ijk})\mel{v}{\mu_{l}}{0}} {\Delta_{g0,gv}}
\end{equation}
The $A$ term depends upon $\order{Q^0}$ transitions (i.e., Condon approximation), the $B$ term depends upon $\order{Q}$ transitions, and the $C$ term depends on $\order{Q^2}$ transitions. 
The C term is suppressed in the following discussion as it depends on one and two photon forbidden transitions. \cite{Ziegler1988, Neddersen1989, Bonang1992}
Note that some reports obtain $A, B, C$ coefficients through a Herzberg-Teller expansion of the electronic states to explicitly expose vibronic couplings in terms of $\partial H / \partial Q $, where H is the electronic Hamiltonian.\cite{HerzbergTeller1933, Petrov1985, Neddersen1989, Baranov1990}
This approach is not used here as the expansion of $\vec{M}^{ij}$ in normal mode coordinates provides sufficient physical insight into hyper-Raman selection rules.

By contracting over the virtual vibrational states $\ket{n}$, the $A$ and $B$ coefficients, where $B = B_1 + B_2$, are written as
\begin{widetext}
\begin{subequations}\label{ABterms}
\begin{equation}
	\begin{split}
		A_{ijk} = \frac{1}{\hbar^2}\sum_{m,e,v'} M^{ge}_{0,i} 
		M^{em}_{0,j} 
		M^{mg}_{0,k}
		 \langle v | v' \rangle
		 \langle v' | 0 \rangle 
		 \frac{1}{\Delta_{ev', g0}}
		 \\
	\end{split}
\end{equation}
	\begin{equation}
		\begin{split}
			B_{1_{ijk}} &= \frac{1}{\hbar^2} \sum_{m,e,v',z} M^{ge}_{0,i} \langle v | v' \rangle \left(
			 \frac{\partial M^{em}_{j}}{\partial Q_z} M^{mg}_{0,k} \mel{v'}{Q_z}{0} 
			+M^{em}_{0,j} \frac{\partial M^{mg}_{k}}{\partial Q_z} \mel{v'}{Q_z}{0} \right) \frac{1}{\Delta_{ev', g0}}\\
		\end{split}
	\end{equation}
	\begin{equation}
	\begin{split}
			B_{2_{ijk}} = \frac{1}{\hbar^2} \sum_{m,e,v',z} \frac{\partial M^{ge}_{i}}{\partial Q_z} M^{em}_{0,j} 
			M^{mg}_{0,k} \mel{v}{Q_z}{v'} 
			\langle v' | 0 \rangle 
			\frac{1}{\Delta_{ev', g0}}
	\end{split}
	\end{equation}
\end{subequations}
\end{widetext}
where terms such as $\langle a | b \rangle$ and $\mel{a}{Q}{b}$ are Franck-Condon factors and Herzberg-Teller-type integrals, respectively. 
These expressions are valid for both singly and doubly resonant HDFG, but based upon the resonance schemes, further simplifications can be made (\textit{vide infra}).
\subsubsection{Singly Resonant HDFG}
In the case of singly resonant HDFG, $\omega_3$ is also detuned from any resonance, which permits contraction over ${v'}$ as the dependence on $\ket{v'}$ in the resonance denominator can be neglected. 
This limit is usually reached when $\abs{\omega_{ev',g0} - (\omega_2 + \omega_3)} \gg 2\Gamma_{ev',g0}$.
In such a limit, assuming a normal mode basis, we see for singly resonant HDFG that 
\begin{subequations}
\begin{equation}
	A_{ijk} \sim \sum_{v'}\langle v | v' \rangle \langle v' | 0 \rangle = \delta_{v0} = 0
\end{equation}
\begin{equation}
	B_{1_{ijk}} = \frac{1}{\hbar^2} \sum_{m,e,z} M^{ge}_{0,i} \left( 
	\frac{\partial M^{em}_{j}}{\partial Q_z} M^{mg}_{0,k}
	+M^{em}_{0,j} \frac{\partial M^{mg}_{k}}{\partial Q_z} \right) \frac{\mel{v}{Q_z}{0}}{\Delta_{e, g0}}
\end{equation}
\begin{equation}
	B_{2_{ijk}} = \frac{1}{\hbar^2} \sum_{m,e,z} \frac{\partial M^{ge}_{i}}{\partial Q_z} M^{em}_{0,j} 
	M^{mg}_{0,k}  
	\frac{\mel{v}{Q_z}{0}}{\Delta_{e, g0}}
\end{equation}
\end{subequations}
i.e., A$_{ijk}$ vanishes, but B$_{ijk}$ is non-zero, for singly resonant HDFG. 
When $\abs{\omega_{ev',g0} - (\omega_2 + \omega_3)} \gtrsim \Gamma_{ev',g0}$, $\Re(\frac{1}{\Delta_{ev', g0}})$ can allow for small $A$ term contributions (cf. the long tail associated with the $A$ term of DR-HDFG as presented in \autoref{fig:doubres_spec}b).
As such, $A$ usually only vanishes when $\omega_2, \omega_3$ are highly detuned from electronic states. 
In the limit where $A$ vanishes, the singly resonant HDFG susceptibility can be approximated as $\chi^{(3)}_{IJKL} \sim \langle B_{ijk} \mu_l \rangle$, or, while a vibrational mode must be infrared active for HDFG output, vibronic activity is necessary for singly resonant HDFG response.
The breakdown of the Condon approximation in singly resonant HDFG supports the relatively small HDFG $\chi^{(3)}$ values ($\sim$ order of magnitude larger than nonresonant background) for hexane and similar organic solvents. \cite{RN350, RN351, RN353}

\subsubsection{Doubly Resonant HDFG}
The selection rules for DR-HDFG are different. 
DR-HDFG (where a vibrational and vibronic state are resonantly coupled) can provide a tool for measuring vibronic coupling, analogous to fully resonant DFG. \cite{Dick83_1, Shen94}
Many multi-resonant nonlinear spectroscopies have been developed to resolve vibronic coupling in molecular samples. \cite{Carlson1990, Gaynor2017, RN276}
DR-HDFG provides methods to investigate vibronic coupling with only two laser pulses.
Experimentally, DR-HDFG would provide identical information to DR-HSFG, but only DR-HDFG is phasematchable in media with normal indices of refraction, so we focus our attention on DR-HDFG. \cite{RN278}
DR-HDFG output is dependent upon the $A,B$ terms prior to contraction over $\ket{v'}$ as written in \autoref{ABterms}.
